# (PART\*) Part A - Demand/Traffic Development <br> during COVID19 {#partA -}

# COVID Briefing {- #covidh1}

The present section aims to compare the impact of COVID-19 crisis on air traffic in Brazil and Europe

```{r}
#Load relevant packages
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(maps)
library(patchwork)
library(flextable)
library(lubridate)
library(zoo)

#Read relevant data
bra_count_airport <- read_csv("./data/PBWG-BRA-airport-traffic.csv") %>%
  mutate(APT_ICAO = as.factor(APT_ICAO))
eur_count_airport <- read_csv("./data/PBWG-EUR-airport-traffic.csv") %>% mutate(APT_ICAO = as.factor(ICAO), .before = DATE, .keep = "unused")

```

# Scope {-}

This briefing includes 2019 and 2020 data from the following airports:
```{r}
#useful filters
bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
eur_apts <- c("EDDF", "EDDM", "EGLL", "EHAM", "LEMD", "LFPG", "LIRF", "LSZH")

scope_function <- . %>% filter(year(DATE) %in% 2019:2020 & APT_ICAO %in% c(bra_10_apts, eur_apts))

bra_count_airport <- bra_count_airport %>% scope_function()
eur_count_airport <- eur_count_airport %>% scope_function()
#head(bra_count_airport)
#head(eur_count_airport)
```

## Brazil {-}

```{r, include=FALSE}
print(unique(bra_count_airport$APT_ICAO))
```

SBBR, SBCF, SBCT, SBGL, SBGR, SBKP, SBPA, SBRJ, SBSP, and SBSV

## Europe {-}

```{r, include=FALSE}
unique(eur_count_airport$APT_ICAO)
#####---THERE ARE MISSING AIRPORTS COMPARED TO THE FULL REPORT---#####
```

European airports are EDDF, EDDM, EGLL, EHAM, LEMD, LFPG, LIRF, and LSZH

# Year in Review {-}

The COVID-19 pandemic hit all countries worldwide, but at different moments. In Europe, the first case was registered in January 24, 2020, in France. The first patient in Brazil was officially announced almost one month later in February 26, 2020. In March 11, 2020, the World Health Organization declared the crisis a "global pandemic".

```{r}
## TO DO
#1. year comparison chart
#2. first four months comparisons

#defining reference percentile
ref_pct   <- 0.9

#utility function
daily_reg_count <- . %>% 
  group_by(DATE) %>% 
  summarize(MVTS = sum(ARRS, DEPS),
            MVTS_INT = sum(ARRS-ARRS_DOM, DEPS-DEPS_DOM)) %>%
  mutate(MVTS_NORM = MVTS / quantile(MVTS[year(DATE)==2019], probs = ref_pct),
         MVTS_INT_NORM = MVTS_INT / quantile(MVTS_INT[year(DATE)==2019], probs = ref_pct) )

eur_count_norm <- eur_count_airport %>% daily_reg_count() %>% mutate(REG = "EUR")
#creating eur rolling avgs
eur_count_norm <- eur_count_norm %>% mutate(MVTS_ROLLAVG = rollmean(MVTS, k = 7, fill = NA), MVTS_NORM_ROLLAVG = rollmean(MVTS_NORM, k = 7, fill = NA))

bra_count_norm <- bra_count_airport %>% daily_reg_count() %>% mutate(REG = "BRA")
#creating bra rolling avgs
bra_count_norm <- bra_count_norm %>% mutate(MVTS_ROLLAVG = rollmean(MVTS, k = 7, fill = NA), MVTS_NORM_ROLLAVG = rollmean(MVTS_NORM, k = 7, fill = NA))

counts_norm <- bind_rows(bra_count_norm, eur_count_norm)

ds4 <- counts_norm %>% group_by(REG, YR = as.factor(year(DATE))) %>% summarise(MVTS = sum(MVTS), MVTS_INT = sum(MVTS_INT)) %>% mutate(VAR_TO_2019 = round((MVTS-MVTS[YR==2019])/MVTS[YR==2019], 2))
#ds4

g1 <- ds4 %>% ggplot() + aes(x = REG) +
  geom_col(position = position_dodge()) + aes(y = MVTS, fill = YR) +
  geom_text(position = position_dodge(0.9), vjust = 1.2) + aes(label = MVTS, y = MVTS) +
  scale_y_continuous(labels = scales::number_format(), limits = c(0, 3500000))

g2 <- ds4 %>% ggplot(aes(x = REG)) +
  geom_col(aes(y = VAR_TO_2019), fill = "royalblue") +
  geom_text(data = ds4[ds4$YR == 2020,], aes(y = VAR_TO_2019, label = scales::percent(VAR_TO_2019, accuracy = 1)), nudge_y = 0.1, color = "white" ) +
  theme(axis.title = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())

g1 + inset_element(g2, left = 0.05, bottom = 0.6, right = 0.4, top = 1)

#option 2
ds4 %>% ggplot(aes(x = YR)) +
  geom_line(aes(y = MVTS, group = REG, color = REG)) +
  scale_y_continuous(limits = c(0, 3500000), labels = scales::label_number())

```


## Daily Traffic {-}

```{r timeline_movs, dpi = 600}
##TO DO rolling avg 7 day, raw numbers in alpha 0.3
#counts_norm
counts_norm %>% ggplot(aes(x = DATE)) +
  geom_point(aes(y = MVTS, colour = REG), alpha = 0.2, size = 0.1) +
  geom_line(aes(y = MVTS_ROLLAVG, colour = REG), size = 1.5) +
  theme_minimal() +
  labs( title    = "PBWG - regional air traffic comparison"
       ,subtitle = "total movements"
       ,x = NULL, y = NULL) +
  scale_colour_manual(values = c("red", "blue"))

counts_norm %>% ggplot(aes(x = DATE)) +
  geom_point(aes(y = MVTS_NORM, colour = REG), alpha = 0.2, size = 0.1) +
  geom_line(aes(y = MVTS_NORM_ROLLAVG, colour = REG), size = 1.5) +
  theme_minimal() +
  labs( title    = "PBWG - regional air traffic comparison"
       ,subtitle = "normalised movements, 7-day rolling average"
       ,x = NULL, y = NULL) +
  scale_colour_manual(values = c("red", "blue"))
```

## Impacts on international flows {-}

```{r}
## TODO 
#1.explain the definitions of Internationl for Br and EUR
#2.check the connections BR-EUR, via ANAC database
#3. make an alternative graph with reduced timeline? check



counts_norm %>%
  ggplot(aes(x = DATE)) +
  geom_line(aes(y = MVTS_INT, colour = REG)) +
  theme_minimal() +
  labs( title    = "PBWG - regional air traffic comparison"
       ,subtitle = "total international movements"
       ,x = NULL, y = NULL)


counts_norm %>%
  ggplot(aes(x = DATE)) +
  geom_line(aes(y = MVTS_INT_NORM, colour = REG)) +
  theme_minimal() +
  labs( title    = "PBWG - regional air traffic comparison"
       ,subtitle = "normalised international movements"
       ,x = NULL, y = NULL)

```

## Impacts on airport level {-}

```{r}
#-----TO DO: AIRPORT BREAKDOWN
#utility function
annual_count <- . %>% group_by(APT_ICAO, YR = year(DATE)) %>% summarise(MVTS = sum(ARRS, DEPS), MVTS_INT = sum(ARRS - ARRS_DOM, DEPS - DEPS_DOM))

multiple_counts <- . %>% mutate(INT_SHARE = round(MVTS_INT / MVTS, 2), VAR_TO_2019 = round((MVTS-MVTS[YR==2019])/MVTS[YR==2019], 2))

ds1 <- eur_count_airport %>% annual_count() %>% multiple_counts() %>% mutate(REG = "EUR")
ds2 <- bra_count_airport %>% annual_count() %>% multiple_counts() %>% mutate(REG = "BRA")
ds3 <- bind_rows(ds1, ds2)
#ds3

ds3 %>% ggplot(aes(x = reorder(APT_ICAO, MVTS))) +
  geom_col(aes(y = MVTS, fill = factor(YR)), position = "dodge") +
  facet_grid(.~REG, scales = "free_x") +
  scale_y_continuous(labels = scales::label_number()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), axis.ticks = element_blank())

ds3 %>% filter(YR == 2020) %>%
  ggplot(aes(x = reorder(APT_ICAO, VAR_TO_2019))) +
  geom_col(aes(y = VAR_TO_2019, fill = REG)) +
  geom_text(aes(y = VAR_TO_2019, label = scales::percent(VAR_TO_2019, accuracy = 1)), size = 2.5, nudge_y = 0.04) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(-0.8,0)) +
  coord_flip()

```
TODO - same, but for jan-apr21 / jan-apr2019
